A semantics of core AUTOSAR

### 2016-08-29 paper/paper.tex
Addressed Patrik's comments.

The process terms of our calculus are as follows:
\begin{itemize}
\item $\Trunnable {\Tr r i} t \Act n$  \newline
Dynamic state for runnable $r$ of component $i$, indicating $n$ current instances, $t$ seconds left until next possible instantiation, and activation state $\Act$.
\item $\Trinst {\Tr r i} c \Xs \Code$ \newline
An instance of runnable $r$ of component $i$, currently executing $\Code$ and owning the exclusive areas $\Xs$, possibly running on behalf of client $c$.
\item $\Texcl {\Tx x i} u$ \newline
Exclusive area $x$ of component $i$, with boolean occupation state $u$.
\item $\Tirv {\Ts s i} v$ \newline
Inter-runnable variable $s$ of component $i$, currently holding value $v$.
\item $\Tdelem {\Tep e p i} u v$ \newline
Non-queued sender-receiver data element $e$ in port $p$ of component $i$, currently holding value $v$ with boolean update status flag $u$.
\item $\Tqelem {\Tep e p i} n \Vs$ \newline
Queued sender-receiver data element $e$ in port $p$ of component $i$ with maximum capacity $n$, currently holding value sequence $\Vs$.
\item $\Toper {\Top o p i} m \Srv$ \newline
Client-server operation $o$ in port $p$ of component $i$, currently holding sequence number $m$ and call state $\Srv$.
\item $\Ttimer {\Tr r i} {t_p} t$ \newline
Periodic timer for runnable $r$ of component $i$, with a period of $t_p$ and $t$ seconds left of its current cycle.
\end{itemize}

@@

\[
\begin{array}{rcll}
  \V{P},\V{Q}
        & ::= & \Trunnable a t {\Act} n      \\
        & |   & \Trinst a c {\Xs} {\Code}    \\
        & |   & \Texcl a u                   \\
        & |   & \Tirv a v                    \\
        & |   & \Tdelem a u v                \\
        & |   & \Tqelem a n {\Vs}            \\
        & |   & \Toper a m \Srv              \\
        & |   & \Ttimer a {\V{T_p}} t        \\
        & |   & \Tpar p q                    \\
        & |   & \Tzero                       \\
\end{array}
\]

@@

TODO: Perhaps match up the order of introducing the terms of the initial process term \V{P_0} with these subsections

@@

TODO: Dangling pointer to Section~\ref{sect:DiscExcl}

@@

Axioms (13) and (15) define unbuffered inter-process communication from a runnable instance point of view. These transitions are just syntactic variants of shared variable reading and writing (axioms (5) and (7)). A data element (\textbf{delem}) process term contains two pieces of state in addition to its address: a communicated data value and a boolean flag for keeping track of unread writes. Axioms (14) and (16) capture reads and writes to the data value and also set the update flag accordingly. The side-condition in axiom (16) makes the rule apply only if there is a static connection from the data element being written and the address of the matched term (recall that data element storage is associated with the receiving side of a connection). Because AUTOSAR allows sender-receiver communication patterns to be one-to-many (such as $\Tconnect {i.p.e} a$ and $\Tconnect {i.p.e} b$), parallel data element updates like the following are possible:
$$
\begin{array}{c}
  \Trinst {i.r} c \Xs {\TrteWrite {p.e} v k} \,\Opar  \\
  {\Tdelem a \_ \_} \Opar {\Tdelem b \_ \_} \vspace{0.6ex} \\
  \red{\say{i.p.e}{\Twr v}} \\
  \Trinst {i.r} c \Xs {\Tap k \Tvoid} \,\Opar \\
  {\Tdelem a \Ttrue v} \Opar {\Tdelem b \Ttrue v}
\end{array}
$$
As an aside, note that the associativity of the parallel composition operator is made entirely irrelevant by the symmetric shape of transition rules (1) and (2).

TODO: ? I don't get that aside ? (associativity is about equality of terms: (a|b)|c == a|(b|c) but the rules deal with transitions, not equality.)

@@

%% TODO: PaJa: continue reading from here.

@@

The AUTOSAR specification contains a very subtle sensitivity to this kind of non-deterministism that is worth examplifying. Consider a queued data element $a$ together with a runnable $b$ set up to trigger on data reception events for $a$.
%**TODO: The explanation is rather long so it may be good to indicate in a few words what it will show already at start: "... lead to the runnable triggering once or twice"
A scenario where two values $v_1$ and $v_2$ are sent on a port connected to $a$ may lead to the following trace:
%
TODO: The i.p.e in the transition label is not explained.
$$
\begin{array}{c}
  {\Tqelem a 2 \Teps} \Opar
  {\Trunnable b 0 \Tidle 0} \vspace{0.6ex} \\

@@

  {\Tqelem a 2 {\Tcons{v_1}{v_2}}} \Opar
  {\Trunnable b 0 \Tidle 1} \,\Opar  \\
  {\Trinst b \Tvoid \Teps {\Code}}
\end{array}
$$
Here there is only one $b$ instance created and none is pending, even though $a$ correctly contains the two values received. The reason is that by swapping the order of the first instantiation and the second data reception steps, the system is put in a state where its inability to count pending instantiations becomes fatal.
%*TODO: Perhaps abbreviate the terms to make the difference more clear: let Q(vs) = \Tqelem a 2 {vs}; R(n) = \Trunnable b 0 \Tidle n; RI = \Trinst b \Tvoid \Teps {\Code}

%   Q(\Teps) \Opar R(0)  ->*  Q(\Tcons{v_1}{v_2}}) \Opar R(1) \Opar RI
% and
%   Q(\Teps) \Opar R(0)  ->*  Q(\Tcons{v_1}{v_2}}) \Opar R(2) \Opar RI \Opar RI

Unfortunately there is very little a programmer can do to influence which of these two traces is taken. Giving $b$ a positive \emph{minimumStartInterval} just guarantees that any subsequent events are lost during that interval. Long delays between reception events will reduce the likelihood of losses, but the choice of when to take an instantiation step is still fully internal to the operating system scheduler. Moreover, when messages are produced by multiple parallel senders, maintaining control over their time distribution may be highly impractical. Non-concurrent runnables can work around the problem by iterating over all received data at each invocation, but that solution is not very attractive if concurrent processing of each received item is desired.

One could argue that replacing the {\bf pending}/{\bf idle} flag of runnables with a small counter variable would render the event triggering mechanism in AUTOSAR much more robust. The standard explicitly excludes event counting for all events but operation invocation, though, so the possibility that events may be lost is undeniably intentional \cite[ch.~4.2.3]{AR:RTE}. However, should this design decision ever be revisited, the two trace alternatives above could serve as a succinct problem characterization.

@@

Unqueued data elements that are read before first written should return the error code {\bf neverReceived} \cite[ch.~5.6.10]{AR:RTE}. Since no exceptions to this rule are mentioned in the specification, our intepretation is that this holds even when a data element possesses an \emph{initValue} attribute  (initial {\bf delem} state in Section~\ref{sec:Sem}). However, as this renders any such initial value assignments entirely redundant, an alternative interpretation would be to create a term
$$
  \Tdelem {i.p.e} \Tfalse v
$$
for each unqueued data element $i.p.e$, where $v$ is the \emph{initValue} of $e$ if it exists, and $\TneverReceived$ otherwise.
%*TODO: perhaps replace i.p.e by a?

@@

Closely related to the interleaving of timing events, and of interest even if only distinct runnables are being triggered, is the notion of initial timer offsets. That is, at what time will each timer tick for the first time --- immediately at startup or at some other (later) time? The AUTOSAR specification only introduces the notion of timing offset in relation to the operating system tasks that are expected to implement runnables; the Software Component Template itself is totally silent on the issue \cite{AR:SWC}. Lacking an abstract definition of runnable offsets, we have chosen to simply fire all timers simultaneously at $0$ seconds from startup (hence the value $0$ in the initial {\bf timer} term definition in Section~\ref{sec:Sem}). It is not hard to see that other values may be equally reasonable, though, like
$$
	\Ttimer {i.r} {t_p} {t_p}
$$
which lets each timer wait one full cycle before triggering for the first time. However, should we really want to take the assumed implementation offsets into account, the initial timer state would rather be
$$
	\Ttimer {i.r} {t_p} {t_o}
$$
for some arbitrarily chosen non-negative $t_o$. One might even want to add the condition $t_o < t_p$, but it must be emphasized that all these initilization options are speculative --- the AUTOSAR Software Component Template must be clarified before a sufficiently abstract model of timed behavior can be obtained.

@@

First the code syntax needs to be extended so that failing computations can be represented:
$$
\begin{array}{lll}
  \Code & ::= & \ldots               \\
        & |   & \Return v            \\
        & |   & \Throw
\end{array}
$$
A "throw" command has no continuation and is thus similar to the exception generating commands of many modern programming languages. It could be given an informative value parameter as well, but this will not be needed in our simplistic approach. Throw commands can appear explicitly in the code to indicate software-detected errors, or they can be the result of special failure transitions, like the following alternative handling of an exit order violation:
$$
\begin{array}{lcll}
		{\Trinst {i.r} c {\Tcons y \Xs} {\TrteExit x k}}
		\;\; \red {\Tsay {i.r} \Tfail} \\
		{\Trinst {i.r} c {\Tcons y \Xs} \Throw}
		& \Pif\;x \neq y
\\ \\
		{\Trinst {i.r} c \Teps {\TrteExit x k}}
		\;\; \red {\Tsay {i.r} \Tfail}  \\
		{\Trinst {i.r} c \Teps \Throw}
\end{array}
$$
Because a failed runnable instance now has an explicit representation, we can specify details of how such a process should interact with its enviroment. We may for example wish to define that a special error code is returned if our runnable is running on behalf of a client, and that any exclusive areas held are automatically exited in the right order (c.f.~axioms (42) and (11)):
$$
\begin{array}{lcll}
		{\Trinst a {\Tclient b m \_} \Xs \Throw}
		\;\; \red{\Tsay b {\Tret m {\kw{serverErr}}}} \\
		{\Trinst a \Tvoid \Xs \Throw}
\\ \\
		{\Trinst {i.r} \Tvoid {\Tcons x \Xs} \Throw}
		\;\; \red{\Tsay {i.x} \Tex} \\
		{\Trinst {i.r} \Tvoid \Xs \Throw}
\end{array}
$$
Finally, we can ensure that the instance counter of a runnable is properly decremented even if an instance terminates due to a failure (in interaction with axiom (52)):
$$
\begin{array}{rcll}
		{\Trinst a \Tvoid \Teps \Throw}
		& \red{\Tsay a \Tterm}
		\Tzero
\end{array}
$$
Other design choices are of course perfectly possible, and more elaborate rules, for example defining an explicit exception-catching mechanism, could be considered as well. We must emphasize, though, that the rules in this section are just hypothetical examples; they are not based on the current AUTOSAR specification and are only presented to illustrate the succinctness and level of detail that can be obtained by chosing a formal semantics aproach.

@@

All aspects of the AUTOSAR specification that concern the representation and selection of data values are beyond the scope of the present study. This includes
\begin{itemize}
\item Data types on any level of abstraction, and their associated mappings.
\item Port interface mapping and data scaling.
\item Measurements and calibration parameters.
\item Variant handling.
\end{itemize}
These modelling elements only impose restrictions and demands on the values $v$ and continuations $k$ we reference in our semantics, and which we in turn do not constrain in any way.

@@

%*TODO: Perhaps avoid repeating the common suffix SwComponentType (with explanation).
The only atomic software component type we include on our formalization is the \emph{ApplicationSwComponentType}. Other atomic component types describe dynamic behaviors that are either just minor variants of the formalized component type (\emph{SensorActuatorSwComponentType}, \emph{ComplexDeviceDriverSwComponentType}, \emph{ServiceSwComponentType}, \emph{ServiceProxySwComponentType}), or trivial in comparison (\emph{NvBlockSwComponentType}, \emph{EcuAbstractionSwComponentType}, \emph{ParameterSwComponentType}).

@@

A few areas of the AUTOSAR SWC and RTE specifications have been omitted for the purpose of keeping the size of the resulting formalism at a reasonable level. These are areas where a worked out formalization could prove interesting, and possibly even discover new design ambiguities. However, we also believe that the techniques required for undertaking such an effort are already demonstrated by our current work, and that making the resulting semantics accessible is an even more important goal than striving for completeness. The major omissions of this kind are:
\begin{itemize}
\item Implicit communication, where designated port elements are automatically read/written at the start/end the of runnables that access them.
\item Category 2 runnables, which declare certain read or receive operations as \emph{WaitPoints} that block progress until the associated events occur. While this form of event handling is fundamentally different from the standard mechansim of activating runnables, it may be noted that the result command that follows from calling a \emph{synchronousServerCallPoint} already defines the essence of such a blocking \emph{WaitPoint}.
\item The AUTOSAR COM layer and related events and commands (\emph{DataWriteCompletedEvent}, \emph{DataSendCompletedEvent}, \emph{DataReceiveErrorEvent}, the feedback command) \cite{AR:COM}. COM is the communication substrate between nodes in a distributed AUTOSAR system and its limited storage capacity introduces an element of resource competition among communicating components. It also leads to the separation of sending and reception into distinct events, as well as a new class of communication errors.
\item Most details found in the \emph{ComSpec} annotations that can be attached to ports (currently only \emph{initValues} are acknowledged).
\item The meta-programming oriented RTE Lifecycle and Callback APIs.
\end{itemize}

@@

%TODO: RuleLong?
\subsection{Calling a server}

>>>>

The process terms of our calculus are as follows:
\begin{itemize}
\item $\Trunnable {\Tr r i} t \Act n$  \newline
Dynamic state for runnable $r$ of component $i$, indicating $n$ current instances, $t$ seconds left until next possible instantiation, and activation state $\Act$.
\item $\Trinst {\Tr r i} c \Xs \Code$ \newline
An instance of runnable $r$ of component $i$, currently executing $\Code$ and owning the exclusive areas $\Xs$, possibly running on behalf of client $c$.
\item $\Tirv {\Ts s i} v$ \newline
Inter-runnable variable $s$ of component $i$, currently holding value $v$.
\item $\Texcl {\Tx x i} u$ \newline
Exclusive area $x$ of component $i$, with boolean occupation state $u$.
\item $\Tdelem {\Tep e p i} u v$ \newline
Non-queued sender-receiver data element $e$ in port $p$ of component $i$, currently holding value $v$ with boolean update status flag $u$.
\item $\Tqelem {\Tep e p i} n \Vs$ \newline
Queued sender-receiver data element $e$ in port $p$ of component $i$ with maximum capacity $n$, currently holding value sequence $\Vs$.
\item $\Toper {\Top o p i} m \Srv$ \newline
Client-server operation $o$ in port $p$ of component $i$, currently holding sequence number $m$ and call state $\Srv$.
\item $\Ttimer {\Tr r i} {t_p} t$ \newline
Periodic timer for runnable $r$ of component $i$, with a period of $t_p$ and $t$ seconds left of its current cycle.
\end{itemize}

@@

\[
\begin{array}{rcll}
  \V{P},\V{Q}
        & ::= & \Trunnable a t {\Act} n      \\
        & |   & \Trinst a c {\Xs} {\Code}    \\
        & |   & \Tirv a v                    \\
        & |   & \Texcl a u                   \\
        & |   & \Tdelem a u v                \\
        & |   & \Tqelem a n {\Vs}            \\
        & |   & \Toper a m \Srv              \\
        & |   & \Ttimer a {\V{T_p}} t        \\
        & |   & \Tpar p q                    \\
        & |   & \Tzero                       \\
\end{array}
\]

@@

@@

@@

Axioms (13) and (15) define unbuffered inter-process communication from a runnable instance point of view. These transitions are just syntactic variants of shared variable reading and writing (axioms (5) and (7)). A data element (\textbf{delem}) process term contains two pieces of state in addition to its address: a communicated data value and a boolean flag for keeping track of unread writes. Axioms (14) and (16) capture reads and writes to the data value and also set the update flag accordingly. The side-condition in axiom (16) makes the rule apply only if there is a static connection from the data element being written and the address of the matched term (recall that data element storage is associated with the receiving side of a connection). Because AUTOSAR allows sender-receiver communication patterns to be one-to-many (such as $\Tconnect {i.p.e} a$ and $\Tconnect {i.p.e} b$), parallel data element updates like the following are possible:
$$
\begin{array}{c}
  \Trinst {i.r} c \Xs {\TrteWrite {p.e} v k} \,\Opar  \\
  {\Tdelem a \_ \_} \Opar {\Tdelem b \_ \_} \vspace{0.6ex} \\
  \red{\say{i.p.e}{\Twr v}} \\
  \Trinst {i.r} c \Xs {\Tap k \Tvoid} \,\Opar \\
  {\Tdelem a \Ttrue v} \Opar {\Tdelem b \Ttrue v}
\end{array}
$$
As an aside, note that the ordering of terms in a parallel composition is made entirely irrelevant by the symmetric formulation of transition rules (1) and (2).

@@

@@

The AUTOSAR specification contains a very subtle sensitivity to this kind of non-deterministism that is worth examplifying. Consider a queued data element $a$ together with a runnable $b$ set up to trigger on data reception events for $a$. A scenario where two values $v_1$ and $v_2$ are sent on a port element $i.p.e$ connected to $a$ may lead to the following trace:
$$
\begin{array}{c}
  {\Tqelem a 2 \Teps} \Opar
  {\Trunnable b 0 \Tidle 0} \vspace{0.6ex} \\

@@

  {\Tqelem a 2 {\Tcons{v_1}{v_2}}} \Opar
  {\Trunnable b 0 \Tidle 1} \,\Opar  \\
  {\Trinst b \Tvoid \Teps {\Code}}
\end{array}
$$
Here there is only one $b$ instance created and none is pending, even though $a$ correctly contains the two values received. The reason is that by swapping the order of the first instantiation and the second data reception steps, the system is put in a state where its inability to count pending instantiations becomes fatal.

So, in an abbreviated form, both the following derivations are possible:
$$
\begin{array}{l}
  q \Opar p \rightarrow \cdots \rightarrow q_1 \Opar p_1 \Opar p_{\text{inst}} \\
  q \Opar p \rightarrow \cdots \rightarrow q_1 \Opar p_2 \Opar p_{\text{inst}} \Opar p_{\text{inst}}
\end{array}
$$
That is, two data reception events in sequence can lead to the creation of either one or two runnable instances, and both behaviors are equally correct according to the AUTOSAR semantics.

Unfortunately there is very little a programmer can do to influence which of these two traces is taken. Giving $b$ a positive \emph{minimumStartInterval} just guarantees that any subsequent events are lost during that interval. Long delays between reception events reduce the likelihood of losses, but the choice of when to take an instantiation step is still fully internal to the operating system scheduler. Moreover, when messages are produced by multiple parallel senders, maintaining control over their time distribution may be highly impractical. Non-concurrent runnables can work around the problem by iterating over all received data at each invocation, but that solution is not very attractive if concurrent processing of each received item is desired.

One could argue that replacing the {\bf pending}/{\bf idle} flag of runnables with a small counter variable would render the AUTOSAR event triggering mechanism much more robust. The standard explicitly excludes counting for all events but operation invocation, though, so the possibility of event losses is undeniably intentional \cite[ch.~4.2.3]{AR:RTE}. However, should this design decision ever be revisited, the two trace alternatives above could serve as a succinct problem characterization.

@@

Unqueued data elements that are read before first written should return the error code {\bf neverReceived} \cite[ch.~5.6.10]{AR:RTE}. Since no exceptions to this rule are mentioned in the specification, our intepretation is that this holds even when a data element possesses an \emph{initValue} attribute  (initial {\bf delem} state in Section~\ref{sec:Sem}). However, as this renders any such initial value assignments entirely redundant, an alternative interpretation would be to create a term
$$
  \Tdelem {i.p.e} \Tfalse v
$$
for each unqueued data element $i.p.e$, where $v$ is the \emph{initValue} of $e$ if it exists, and $\TneverReceived$ otherwise.

@@

Closely related to the interleaving of timing events, and of interest even if only distinct runnables are being triggered, is the notion of initial timer offsets. That is, at what time will each timer tick for the first time --- immediately at startup or at some other (later) time? The AUTOSAR specification only introduces the notion of timing offset in relation to the operating system tasks that are expected to implement runnables; the Software Component Template itself is totally silent on the issue \cite{AR:SWC}. Lacking an abstract definition of runnable offsets, we have chosen to simply fire all timers simultaneously at $0$ seconds from startup (hence the value $0$ in the initial {\bf timer} term definition in Section~\ref{sec:Sem}). It is not hard to see that other values may be equally reasonable, though, like
$$
	\Ttimer {i.r} {t_p} {t_p}
$$
which lets each timer wait one full cycle before triggering for the first time. However, should we really want to take the assumed implementation offsets into account, the initial timer state would rather be
$$
	\Ttimer {i.r} {t_p} {t_o}
$$
for some chosen non-negative $t_o$. One might even want to add the condition $t_o < t_p$, but it must be emphasized that all these initilization options are speculative --- the AUTOSAR Software Component Template must be clarified before a sufficiently abstract model of timed behavior can be obtained.

@@

First the code syntax needs to be extended so that failing computations can be represented:
$$
\begin{array}{lll}
  \Code & ::= & \ldots               \\
        & |   & \Return v            \\
        & |   & \Throw
\end{array}
$$
A "throw" command has no continuation and is thus similar to the exception generating commands of many modern programming languages. It could be given an informative value parameter as well, but this will not be needed in this simplistic proposal. Throw commands can appear explicitly in the code to indicate software-detected errors, or they can be the result of special failure transitions, like the following alternative resolution of an exit order violation:
$$
\begin{array}{lcll}
		{\Trinst {i.r} c {\Tcons y \Xs} {\TrteExit x k}}
		\;\; \red {\Tsay {i.r} \Tfail} \\
		{\Trinst {i.r} c {\Tcons y \Xs} \Throw}
		& \Pif\;x \neq y
\\ \\
		{\Trinst {i.r} c \Teps {\TrteExit x k}}
		\;\; \red {\Tsay {i.r} \Tfail}  \\
		{\Trinst {i.r} c \Teps \Throw}
\end{array}
$$
Because a failed runnable instance now has an explicit representation, we can specify details of how such a process should interact with its enviroment. We may for example wish to define that a special error code is returned if our runnable is running on behalf of a client, and that any exclusive areas held are automatically exited in the right order (c.f.~axioms (42) and (11)):
$$
\begin{array}{lcll}
		{\Trinst a {\Tclient b m \_} \Xs \Throw}
		\;\; \red{\Tsay b {\Tret m {\kw{serverErr}}}} \\
		{\Trinst a \Tvoid \Xs \Throw}
\\ \\
		{\Trinst {i.r} \Tvoid {\Tcons x \Xs} \Throw}
		\;\; \red{\Tsay {i.x} \Tex} \\
		{\Trinst {i.r} \Tvoid \Xs \Throw}
\end{array}
$$
Finally, we can ensure that the instance counter of a runnable is properly decremented even if an instance terminates due to a failure (in interaction with axiom (52)):
$$
\begin{array}{rcll}
		{\Trinst a \Tvoid \Teps \Throw}
		& \red{\Tsay a \Tterm}
		\Tzero
\end{array}
$$
Other design choices are of course perfectly possible, and more elaborate rules, for example involving an explicit exception-catching mechanism, could be considered as well. We must emphasize, though, that the rules in this section are just hypothetical examples; they are not based on the current AUTOSAR specification and are only presented to illustrate the succinctness and level of detail that can be obtained by a formal semantics aproach.

@@

All aspects of the AUTOSAR specification that concern representation and selection of data values are beyond the scope of the present study. This includes
\begin{itemize}
\item Data types on any level of abstraction, and their associated mappings.
\item Port interface mapping and data scaling.
\item Measurements and calibration parameters.
\item Variant handling.
\end{itemize}
These modelling elements only impose restrictions and demands on the values $v$ and continuations $k$ we reference in our semantics, and which we in turn do not constrain in any way.

@@

The only atomic software component type we include on our formalization is the \emph{Application} software component type. Other atomic component types describe dynamic behaviors that are either just minor variants of the formalized component type (\emph{SensorActuator}, \emph{ComplexDeviceDriver}, \emph{Service}, \emph{ServiceProxy}), or trivial in comparison (\emph{NvBlock}, \emph{EcuAbstraction}, \emph{Parameter}).

@@

A few areas of the AUTOSAR SWC and RTE specifications have been omitted for the purpose of keeping the size of the resulting formalism at a reasonable level. These are areas where a worked out formalization could prove interesting, and possibly even discover new design ambiguities. However, we also believe that the techniques required for undertaking such an effort are already demonstrated by our current work, and that making the resulting semantics accessible is an even more important goal than striving for completeness. The major omissions of this kind are:
\begin{itemize}
\item Implicit communication, where designated port elements are automatically read/written at the start/end the of runnables that access them.
\item Category 2 runnables, which declare certain read or receive operations as \emph{WaitPoints} that block progress until the associated events occur. While this form of event handling is fundamentally different from the standard mechansim of activating runnables, it may be noted that the result command that follows from calling a \emph{synchronousServerCallPoint} already defines the essence of such a blocking \emph{WaitPoint}.
\item The AUTOSAR COM layer and related events and commands (\emph{DataWriteCompleted}, \emph{DataSendCompleted}, \emph{DataReceiveError}, the {\em Feedback} command) \cite{AR:COM}. COM is the communication substrate between nodes in a distributed AUTOSAR system and its limited storage capacity introduces an element of resource competition among communicating components. It also leads to the separation of sending and reception into distinct events, as well as a new class of communication errors.
\item Most details found in the \emph{ComSpec} annotations that can be attached to ports (currently only \emph{initValues} are acknowledged).
\item The meta-programming oriented RTE {\em Lifecycle} and {\em Callback} APIs.
\end{itemize}

@@

\subsection{Calling a server}
